Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-05 _tex/index.tex
Built site for gh-pages

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) and its associated
structural causal models (SCM) to provide a graphical and formal
representation of the causal structure of a system \citep{Neal_2020}. In
this context, a \emph{graph} denotes a collection of nodes connected by
edges, where nodes represent random variables. The term \emph{directed}
indicates that the edges of the graph extend from one node to another,
with arrows showing the direction of causal influence. Moreover, the
term \emph{acyclic} indicates the causal influences do not form a loop,
meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

One key advantage of DAGs is that they can represent various causal
structures using only five fundamental building blocks, regardless of
complexity. The left panels of Figure~\ref{fig-dags_scms} illustrates
the \emph{simplified} graphs of these block. In these graphs, the
variables of interest or \emph{endogenous} variables, \(V=\{X,Z,Y\}\),
are depicted as solid black circles, indicating that they are observed
variables. The arrows in the graph reflect the expected direction of
causal influences among the variables.

In contrast, the middle panels of Figure~\ref{fig-dags_scms} depicts the
\emph{magnified} graphs of these building blocks. These graph include,
in addition to the endogenous variables, the \emph{exogenous} variables
\(E=\{e_{X},e_{Z},e_{Y}\}\). These exogenous variables, commonly
referred to as \emph{disturbances} or \emph{errors}, represent factors
not explicitly modeled. Moreover, they are depicted as open circles,
indicating their unobserved nature. These exogenous variables are often
omitted for simplicity, resulting in the simplified DAGs shown in the
left panels. However, including them in the graph can be advantageous in
certain scenarios, as they can help to highlight potential issues
related to conditioning and confounding \citep{Cinelli_et_al_2020},
concepts introduced in the next section.

@@

\caption{\label{fig-dags_scms}The five fundamental building blocks of
DAGs with corresponding SCMs. \textbf{Note:} left column shows
simplified DAGS, middle column show magnified DAGs, and right column
shows the corresponding SCMs.}

@@

A careful examination of the left and middle panels of
Figure~\ref{fig-dags_scms} reveal what do the building blocks assume.
Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1} depict two unconnected
nodes, representing an scenario where variables \(X\) and \(Y\) are not
causally related. Figures \ref{fig-sdag_bb2} and \ref{fig-mdag_bb2} show
two connected nodes, illustrating a scenario where a parent node \(X\)
exerts a causal influence on a child node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-sdag_bb3} and
\ref{fig-mdag_bb3} depict a \emph{chain} (or \emph{pipe}), where \(X\)
influences \(Z\), and \(Z\) influences \(Y\). In this configuration,
\(X\) is a parent node of \(Z\), and \(Z\) a parent node of \(Y\).
Furthermore, the DAG shows that \(X\) is an \emph{ancestor} of \(Y\) and
that the relationship between these variables is entirely
\emph{mediated} by \(Z\). Figures \ref{fig-sdag_bb4} and
\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
\(Y\) are both influenced by \(Z\). In this scenario \(Z\) is a parent
node of both \(X\) and \(Y\). Finally, figures \ref{fig-sdag_bb5} and
\ref{fig-mdag_bb5} depict a \emph{collider}, also known as
\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
related to each other but both influence \(Z\).

>>>>

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG), and their associated
structural causal models (SCM), to provide a graphical and formal
representation of the causal structure of a system \citep{Neal_2020}. In
this context, a \emph{graph} denotes a collection of nodes connected by
edges, where nodes represent random variables. The term \emph{directed}
indicates the edges of the graph extend from one node to another, with
arrows showing the direction of causal influence. Moreover, the term
\emph{acyclic} indicates the causal influences do not form a loop,
meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

Regardless of complexity, DAGs can represent various causal structures
using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. The left panels of
Figure~\ref{fig-dags_scms} illustrate these building blocks using the
\emph{standard} representation \citep{Morgan_et_al_2014}. In these
graphs, the variables of interest, or \emph{endogenous} variables
\(V=\{X,Z,Y\}\), are depicted as solid black circles to signify that
they are observed random variables. The arrows in the graphs reflect the
expected direction of causal influences among these variables.

In contrast, the middle panels of Figure~\ref{fig-dags_scms} present the
building blocks using the \emph{magnified} representation. These graphs
incorporate \emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\),
depicted as open circles to signify their unobserved nature. Exogenous
variables represent \emph{errors} or \emph{disturbances} arising from
omitted factors that the investigator chooses not to model explicitly
\citep[27,68]{Pearl_2009}. Thus, they can be viewed as the
`idiosyncratic' causes of each variable. Although these variables are
often omitted for simplicity, as illustrated in the left panels,
including them in the graphs can be beneficial, as their presence can
reveal potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}, concepts explored in the next section.

@@

\caption{\label{fig-dags_scms}The five fundamental building blocks of
DAGs with corresponding SCMs. \textbf{Note:} left panels show the
standard representation of DAGs, middle panels show the magnified
representation of DAGs, and the right panels show their corresponding
SCMs.}

@@

A careful examination of the left and middle panels of
Figure~\ref{fig-dags_scms} reveals the assumption inherent in these
building blocks. Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1}
depict two unconnected nodes, representing a scenario where variables
\(X\) and \(Y\) are not causally related. Figures \ref{fig-sdag_bb2} and
\ref{fig-mdag_bb2} show two connected nodes, illustrating a scenario
where a parent node \(X\) exerts a causal influence on a child node
\(Y\). Consequently, \(Y\) is considered a \emph{descendant} of \(X\).
Figures \ref{fig-sdag_bb3} and \ref{fig-mdag_bb3} depict a \emph{chain}
(or \emph{pipe}), where \(X\) influences \(Z\), and \(Z\) influences
\(Y\). In this configuration, \(X\) is a parent node of \(Z\), and \(Z\)
is a parent node of \(Y\). Furthermore, the DAG shows that \(X\) is an
\emph{ancestor} of \(Y\) and that \(Z\) completely \emph{mediates} the
relationship between these variables. Figures \ref{fig-sdag_bb4} and
\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
\(Y\) are both influenced by \(Z\). In this scenario, \(Z\) is a parent
node of \(X\) and \(Y\). Finally, Figures \ref{fig-sdag_bb5} and
\ref{fig-mdag_bb5} depict a \emph{collider}, also known as
\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
related to each other but both influence \(Z\).
